Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-01-03 text/sections/00-abstract.tex
Updates.
writing revision (working on the final revision).

Public universities employ more lecturers and fewer professors than at any other point in the last thirty years, relative to student enrolment.
At the same time, state funding for higher education has stagnated.
This paper shows falling state funding caused a substitution away from professors towards lecturers at US public universities, using a shift-share approach to instrument for state funding cuts relative to enrolment.
Universities employ $4.4$\% more lecturers per student following a $10$\% funding cut, relying less on professors by employing $1.4$\% fewer assistant professors and $1.2$\% fewer full professors per student.
Incumbent professors' salaries, promotion rate, and quit rate at Illinois universities are not affected by funding cuts, so that the substitution arose by limiting hiring of new tenure-track/tenured professors.
Stagnating public funding affects faculty and public universities as institutions, likely contributing to worsening student outcomes at public universities since the 1990s.

\vfill
\noindent
\textbf{Key-words:}
State and Local Budget and Expenditures,
Higher Education,
Public Sector Labour Markets

>>>>

Public universities employ more lecturers and fewer professors than at any other point in the last thirty years, relative to student enrolment.
At the same time, state funding for higher education has stagnated.
This paper shows that the decline in state funding led to a substitution away from professors toward lecturers at US public universities.
Using a shift-share approach to instrument for state funding, I find that universities employ 4.4\% more lecturers per student following a 10\% funding cut.
This shift is accompanied by a reduction in assistant professors and full professors per student by 1.4\% and 1.2\%, respectively.
Incumbent professors' salaries, promotion rates, and quit rates at Illinois universities remain unaffected by funding cuts, indicating that the substitution arose from limiting the hiring of new tenure-track/tenured professors.
Stagnating state funding impacts public universities and faculty, likely contributing to the deterioration of student outcomes at public universities since the 1990s.

\vfill
\noindent
\textbf{Key-words:}
Faculty,
Higher Education,
State Funding

### 2024-01-03 text/sections/01-introduction.tex
Updates.
writing revision (working on the final revision).

@@

I use two different analyses to explore the causal effects of state funding on faculty composition and salaries.
My first analysis uses data from the \cite{ipeds} to study the impact of changes in state funding for all US public universities 1990--2021.
My second analysis uses panel data for every faculty member at Illinois public universities for 2010--2021 \citep{ibhed}, to measure how incumbent faculty (i.e., those already employed at the university) are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 per student (or a 10\% increase), do they change how many faculty they employ?
If so, which positions do universities substitute away from, or towards, and do these funding cuts affect the faculty themselves?

@@

In my national-level analysis (IPEDS), I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards lecturers.
A funding cut of \$1,000 per student leads to an average university employing 6 more lecturers;
in percentage terms, a funding cut of 10\% per student leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for many years after the initial funding shock, showing that the effect is not isolated to the year of the funding cut. 
Over the same time period, state funding per student fell by around 35\%, professors per student fell by 9\%, and lecturers per student increased by 99\%.
These results show that falls in state funding explain around 53\% of the fall in professors per student,
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99
In my analysis of all Illinois faculty (IBHED), I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by state funding cuts.
This implies that faculty substitution arose by limiting the hiring of new professors, which is supported by suggestive evidence on cumulative faculty hiring counts.
The substitution is persistent for up to ten years, meaning that universities did not resume professor hiring in years following the funding cut, making the substitution towards lecturers long-run.

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
In addition, empirical papers measure how universities responded to endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects on faculty composition or salaries, or other outcomes related to university instruction.

Another contribution of my paper is to provide evidence on possible mechanisms connecting state funding cuts for higher education and worsening student outcomes.
Substitution towards lecturers is likely a cost-cutting measure for funding constrained universities, though relying on these faculty --- who are often over-worked and not granted long-term employment protections --- may lead to worse student outcomes \citep{ehrenberg2005tenured,zhu2021limited,jaeger2011examining}.
My results provide evidence that the long term trends in higher education funding are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

This paper proceeds as follows.
\autoref{sec:data} describes the data for university finances and faculty in Illinois, and trends in public university funding for the last three decades.
\autoref{sec:conceptual} gives the conceptual framework for how state funding may affect faculty.
\autoref{sec:empirics} draws the empirical framework for isolating the causal effects of state funding on faculty composition and individual faculty, with \autoref{sec:results} presenting the empirical results.
\autoref{sec:discussion} discusses the context and implications for the findings.
\autoref{sec:conclusion} concludes.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\section{Introduction}
\label{sec:intro}

@@

I use two different analyses to explore the causal effects of state funding on faculty composition and salaries.
My first analysis uses data from the \cite{ipeds} to study the impact of changes in state funding for all US public universities 1990--2021.
My second analysis uses panel data for every faculty member at Illinois public universities for 2010--2021 \citep{ibhed}, to measure how incumbent faculty (i.e., those already employed at the university) are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 per student (or a 10\% increase), do they change how many faculty members they employ?
If so, which positions do universities substitute away from, and do these funding cuts affect the faculty themselves?

@@

In my university-level analysis (IPEDS), I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards lecturers.
A funding cut of \$1,000 per student leads to an average university employing 6 more lecturers;
in percentage terms, a funding cut of 10\% per student leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for many years after the initial funding shock, showing that the effect is not isolated to the year of the funding cut. 
Over the same time period, state funding per student fell by around 35\%, professors per student fell by 9\%, and lecturers per student increased by 99\%.
These results show that falls in state funding explain around 53\% of the fall in professors per student,
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99
In my analysis of all Illinois faculty (IBHED), I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by state funding cuts.
This implies that faculty substitution arose by limiting the hiring of new professors, and is supported by suggestive evidence on professor hiring counts.
The substitution is persistent for up to ten years, meaning that universities did not resume professor hiring in years following the funding cut, making the substitution towards lecturers long-run.

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
In addition, empirical papers measure how universities responded to endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects on faculty composition or salaries, or other outcomes related to university instruction.
My results provide evidence that the long term trends in higher education funding and faculty outcomes are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

Another contribution of my paper is to provide evidence on possible mechanisms connecting state funding cuts for higher education and worsening student outcomes.
Substitution towards lecturers is likely a cost-cutting measure for funding constrained universities, but universities that rely on more lecturers have persistently lower student retention graduation rates \citep{jaeger2011examining,ehrenberg2005tenured}.\footnote{
    \cite{bettinger2010does} provide an alternative perspective, that adjunct lecturers improve subject retention for undergraduates, but do not consider graduation rate or any other student outcomes.
    \cite{figlio2015tenure} have similar results at a highly-selective private university, a different sector to that studied in this paper.
}
According to survey evidence, around 50\% of lecturers are employed full-time, with a guaranteed annual salary and university benefits, and the remaining 50\% are employed part-time on adjunct contracts, without university benefits and paid by the class \citep{aau2014Characteristics}.
Adjunct lecturers consistently lead to worse students outcomes, compared to part-time lecturers, largely thanks to the worse working conditions for adjunct lecturers \citep{zhu2021limited}.
Combining the evidence that adjunct lecturers lead to worse student outcomes with the trends in education finances and performance \citep{NBERw23736,NBERw27885}, implies that substituting towards lecturers contributed to deteriorating student outcomes at public universities \citep{NBERw23736,NBERw27885}.

This paper proceeds as follows.
\autoref{sec:data} describes the data for university finances and faculty in Illinois, and trends in public university funding for the last three decades.
\autoref{sec:conceptual} gives the conceptual framework for how state funding may affect faculty.
\autoref{sec:empirics} draws the empirical framework for isolating the causal effects of state funding on faculty, with \autoref{sec:results} presenting the empirical results.
\autoref{sec:discussion} discusses the context and implications for the findings.
\autoref{sec:conclusion} concludes.

### 2024-01-03 text/sections/02-data.tex
writing revision (working on the final revision).

\subsection{Data Description}
The data used in this paper come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information including finances and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e., 1990 refers to the academic year that ran August 1989 to July 1990.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990--2017, and manually combine raw NCES data on 2018-2021 for all relevant variables.
    Figures for enrolment and faculty counts come from the raw NCES version of IPEDS for all years, addressing inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022}, so that these institutions are not considered.

IPEDS reports the count of professors employed by position, as well as total salary expenditures by position.
This gives a resulting panel data-set, where each row represents a university-year, and includes columns for university finances, plus total count and average salary\footnote{
    Real salary is computed by scaling nominal salary to 2021 dollars by the CPI-U.
} for professors by position (lecturer, assistant, tenured, total).
\autoref{tab:ipeds-summary} presents summary statistics for relevant variables in these data.

IPEDS provides information at the university level, but only provides aggregated information within a university.\footnote{
    IPEDS additionally gives a mean salary measure for faculty by rank (inconsistently) for each university.
    Unfortunately, this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period compared to summary statistics provided in \citealt{aau2021survey}.
    Similarly, IPEDS reports of the average salary paid to professors at private institutions also disagree with other sources (including those calculated for Illinois in IBHED data later in this section) so are not considered in this analysis.
}
To investigate the distribution, I integrate individual-level data for every public university professor in the state of Illinois between the years 2010-2021.
IBHED hosts the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These publicly available data provides the basis to build a panel of Illinois public university professors 2010-2021; I define a professor as an individual by their first plus last name and university pairing, and link this database to IPEDS regarding finances for their employing institution.
The Illinois sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Additionally, further analysis using the first year of a professor's employment focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.

@@

States vary largely in how much they fund their public university systems, thanks in part to the state-wide budgeting process.
Planning for an annual budget begins two years ahead of the fiscal year, and the legislature votes to approve or reject the governor office's budget request a number of months before the fiscal year begins.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state appropriations, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
Public universities have relatively lower lobbying power than other funding beneficiaries, so often bear the brunt of state government funding cuts \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
These features lead to yearly variation in state funding not seen in other revenues sources (such as federal funding), since US states differ in their support for higher education, and are subject to fiscal constraints.
Additionally, the number of higher education students in each state varies thanks to the size of each birth cohort.
For example, the birth cohort of 1971 was larger than for neighbouring years, leading to more student demand for limited public university places for students turning 18 in 1989 \citep{bound2007cohort}.
This leads to variation in the amount of funding per student across years, showing that per student state funding varies on multiple dimensions.

@@

Illinois funding for higher education has stagnated, and has experienced serious declines in the decade 2010--2021 --- similar in magnitude to the nation-wide decline over 1990--2021.
State funding among the seven Illinois public university campuses has fallen by over 50\% for the last decade, on both an absolute and per-student basis (see \autoref{fig:illinois-funding}).
Similar to the figures for the rest of the country, there has been a corresponding substitution towards tuition revenue.

>>>>

\subsection{Data Description}
The data used in this paper come from two primary sources: \citet[IPEDS]{ipeds} for data on university funding, finances, and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent for 1990--2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e., 1990 refers to the academic year that ran August 1989 to July 1990.
}
and provide information on university funding, enrolment, and numerous other characteristics.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990--2017, and manually combine raw National Center for Education Statistics (NCES) data on 2018--2021 for all relevant variables.
    Figures for enrolment and faculty counts come from the raw NCES version of IPEDS for all years, addressing inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions have the largest majority of faculty employed and students enrolled.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022}, so that these institutions are not considered.
IPEDS reports the count of professors employed by position.\footnote{
    IPEDS gives a mean salary measure for faculty by rank, but these figures have many missing values and disagree with calculated values from other reputable sources.
    Over 40\% of university--year observations are missing in the IPEDS panel data-set, representing an average of 55\% undergraduate enrolment in each year.
    Yearly averages of faculty salaries by the non-missing values do not agree with trends in average professor salary over the sample time period compared to summary statistics provided in \cite{aau2021survey}.
    IPEDS values consistently report that public university professors are paid on average 20\% more than private sector faculty, which disagrees with other sources on the salary gap direction, and IPEDS figures are consistently higher than averages calculated from individual-level IBHED data.
    For these reasons I do not further analyse IPEDS salary data, and use salary data from IBHED to analyse faculty salaries.
    Notably, IPEDS data on faculty counts do not suffer from these problems.
}
This gives a resulting panel data-set, where each row represents a university-year, and includes columns for university funding and tuition revenue, plus total count\footnote{
    Real salary is computed by scaling nominal salary to 2021 dollars by the CPI-U.
} for each faculty position (lecturer, assistant professor, tenured professor, and total faculty), and other university characteristics.
\autoref{tab:ipeds-summary} presents summary statistics for these variables in IPEDS data.

IPEDS provides information at the university level, but only provides aggregated information within a university.
To investigate outcomes for individual faculty members, I integrate individual-level data for all faculty members at all Illinois public universities 2010--2021.
These data allow me to investigate how individual faculty members are affected by funding cuts to their university, investigating outcomes are more granular details than total faculty counts.
IBHED freely provides this information, as the state of Illinois is required to publicly report base salary and benefits for all administrators, faculty members, and instructors employed by each public college or university.\footnote{
    Public Act 96-0266, effective 1 January 2010, is the relevant Illinois law that requires publicly publishing salary data for all public university faculty salary and benefits \citep{illinois-public-act}.
    A pdf copy of Public Act 096-0266 is included in the supplmentary data of this paper.
    This law applies to all nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These data provide the basis to build a panel of Illinois public university faculty across years 2010--2021; I define a faculty member as an individual by their first plus last name and university pairing,\footnote{
    Some faculty are listed multiple times in these panel data.
    For example, Professor Alberto Agustin Lopez-Scala of the University of Illinois Chicago has two appointments, one as an adjunct faculty and one as a department director.
    For this observation I take the highest paid position (department director) as the primary appointment and drop the secondary appointment (adjunct faculty).
    In analysis of faculty exit rate, I collapse the appointments into one and only consider faculty exits as having a faculty appointment versus no appointment.
}
and link this database to IPEDS for data on their university employer.
The Illinois sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Analysis of professors' in their first year on the job focuses on the subset of professors with observed year of hiring 2011--2021 (i.e., after the panel's first year), representing 1,778 professors in 2011, and 9,099 in 2021.

@@

States vary largely in how much they fund their public university systems, thanks in part to the state-wide budgeting process \citep{NBERw23736}.
Planning for an annual budget begins two years ahead of the fiscal year, and the legislature votes to approve or reject the governor office's budget request a number of months before the fiscal year begins.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state funding, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (e.g., caused by changing economic conditions) necessarily affect state expenditures.
Public universities have lower lobbying power than other state institutions, so often bear the brunt of state government funding cuts \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
These features lead to yearly variation in state funding not seen in other revenues sources (such as federal funding), since US states differ in their support for higher education, and are subject to fiscal constraints.
Additionally, the number of higher education students in each state varies thanks to the size of each birth cohort.
For example, the birth cohort of 1971 was larger than that of 1970 or 1972, leading to more student demand for limited public university places for students turning 18 in 1989 \citep{bound2007cohort}.
This leads to yearly variation in the amount of funding per student across years, showing that per student state funding varies on multiple dimensions.

@@

Illinois funding for higher education has stagnated, and has experienced serious declines in the decade 2010--2021 --- similar in magnitude to the nation-wide decline over 1990--2021.
State funding Illinois public universities fell by over 50\% over 2010--2021, in both absolute and per-student terms (see \autoref{fig:illinois-funding}).
%Similar to the figures for the rest of the country, there has been a corresponding substitution towards tuition revenue.

### 2024-01-03 text/sections/03-concepts.tex
writing revision (working on the final revision).

Universities could respond to funding cuts by cutting faculty salaries, leading to more faculty leaving for jobs at other universities.
Multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, and Cornell University implemented a professor salary cut in 2020.\footnote{
    The Cornell University administration expected large fiscal squeezes in mid-2020, so imposed a faculty salary cut, and then returned the amount cut later in 2020 when the financial squeeze did not materialise.}
Funding cuts could also limit the amount public universities can offer to new faculty hires, leading to fewer or less accomplished faculty accepting these offers.\footnote{
    Faculty often consider their outside options in job decisions, considering offers for employment and/or promotion from multiple universities at the same time (see \citealt{blackaby2005} for an overview of faculty outside options).
    If public universities can only make low salary offers thanks to funding cuts, then they are less likely to land professors with multiple offers.
}
This effect may not be the same across faculty seniority; most universities hire more junior faculty each year, so that yearly hiring disruptions may disrupt the looser market for established full professors more than tighter market for assistant professors or lecturers.
While these factors make it unclear who will be affected most by funding cuts, there is one empirical fact worth noting: the average Illinois assistant professor is paid $\$76,897$, tenured professor $\$109,283$, while lecturers earn $\$31,449$ more than double that of a lecturer (annual salary for 2010--2021 in 2021 USD, see \autoref{tab:illinois-summary}).
Secondly, lecturers are (by definition) granted shorter contracts than professors; assistant professors mostly work on 5--6 year long contracts, full professors on indefinite contracts (tenure), lecturers on yearly or termly contracts.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer resources, substituting away from tenure-track and tenured professors towards lecturers is rationalisable.

It is not clear when the effects of funding cuts will be realised.
University departments often employ faculty on multiple-year contracts (e.g., the tenure contract), and individual professors' decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for university funding to change the faculty composition, or trickle down to faculty.
For example, university departments coordinate hiring in annual cycles, so that a budget short-fall in year $t$ may have no effect on the in-progress hiring committee and resulting hiring decisions, yet the hiring cycle in year $t+1$ may be postponed or cancelled.
As such, this paper investigates the dynamics of state funding shocks' effects on faculty, for the years following a state funding cut.

>>>>

Universities could respond to funding cuts by cutting faculty salaries, leading to more faculty leaving for jobs at other universities.
Multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, and Cornell University implemented a professor salary cut in 2020 at the onset of the Covid-19 pandemic.\footnote{
    The Cornell University administration expected large fiscal squeezes in mid-2020, so imposed a faculty salary cut, and then returned the amount cut later in 2020 when the financial squeeze did not materialise.}
Funding cuts could also limit the salaries public universities can offer to new faculty hires, leading to fewer or less accomplished faculty accepting offers at public universities.\footnote{
    Faculty often consider their outside options in job decisions, considering offers for employment and/or promotion from multiple universities at the same time --- see \cite{blackaby2005} for an overview of faculty outside options.
    If public universities can only make low salary offers thanks to funding cuts, then they are less likely to land professors with multiple offers.
}
This effect may not be the same across faculty seniority; most universities hire more junior faculty each year, so that yearly hiring disruptions may disrupt the looser market for established full professors more than tighter market for assistant professors or lecturers.
While these factors make it unclear who will be affected most by funding cuts, there is one empirical fact worth noting: lecturers have substantially lower annual salaries than professors.
The average Illinois assistant professor's annual salary is $\$76,897$, tenured professors' is $\$109,283$, while lecturers' is $\$31,449$ (annual salary for 2010--2021 in 2021 USD, see \autoref{tab:illinois-summary}).
Secondly, hiring a new professor is a long-term costly investment, when compared to hiring a new lecturer.
Lecturers have shorter employment contracts than professors; assistant professors mostly work on 5--6 year long contracts, full professors on indefinite contracts (tenure), lecturers on yearly or termly contracts.
If a public university's primary obligation is to teach, and they must fulfil this objective with less funding, then substituting away from tenure-track and tenured professors towards lecturers is rationalisable purely on this basis.

It is not clear when the effects of state funding cuts will be realised.
Universities employ professors on multiple-year contracts (e.g., the tenure contract), and decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for effects to trickle down to faculty.
On the other hand, university departments hire in annual cycles, so that a funding cut in one year may immediately impact faculty hiring in the same year, and thus faculty counts in the next academic year.
Yet, it could be the case that faculty hiring resumes the very next year, or even hires more to compensate for last year's cancelled hiring cycle.
This would mean that state funding cuts leads to only transitory changes in faculty counts.
As such, this paper investigates the dynamics of state funding cuts and their effects on faculty, by employing local projections methods to measure the persistence of effects for multiple years following a state funding cut.
